| Early stopping in clinical PET studies: \n| how to reduce expense and exposure

### 2020-10-12 ArticleDraft/Draft.Rmd
Updated draft after pre-print feedback

Figure \ref{fig_k_at_5fpr}A and 3B show the estimated false positive rate when applying sequential BF testing in the case in which there is no difference between the two groups. Three different stopping thresholds have been used (BF > 3, 4 and 6, respectively). When the maximum allowed sample size increases, the false positive rate goes up. This is because the longer the sequential testing can go on, the more decisions are being made. Some of these decisions will be wrong, meaning that the researcher will claim support for an effect, when in fact there is no effect in the population. Eventuelly, when N becomes high enough, the rate of false positives will reach a reach an asymptote (see Supplementary Figure S1). But with a maximum sample size above 30 subjects/group, it is not possible to keep false positives below the 5% using a decision threshold of 4. Instead, if the researchers wish to control the false positives at the 5% level, one option is to increase the decision threshold. Figure \ref{fig_k_at_5fpr}C and 3D show the estimated decision threshold that would be needed to keep the false positive rate at 5% when using sequential testing with the default BF t-test, while still starting at 12 subjects/group and checking the BF after the collection of each patient-control pair.

@@

Figure \ref{fig_powerPhase1} shows the results from the simulations when Nmax=30 subjects/group, using a decision threshold just above 4. In the upper panel the rate of positive studies at different population effects is visualised. The curves cross the y-axis at D=0 (i.e., the false positive rate), 5% for both BF seq and the conventional fixed N test. As can be seen, had the researcher instead applied a fixed N approach (at 30 subjects/group), they would have increased the power to detect the population effect by no more than ~10% for any effect size compared to using sequential testing. Note that when calibrating the BF threshold to correspond to a 5% false positive rate, and stopping only for H~1~, the BF procedure will show identical results to performing sequential NHST t-tests where the significance level has been corrected (see Supplementary Figure SXXX).

@@

Simulation results for different choices of Nmax (15, 20, 50 and 100 subjects/group) are presented in supplementary information (Supplementary Figures S2-S5).

@@

The same results but for different Nmax (15, 20, 50 and 100 subjects/group) are presented in supplementary information (Supplementary Figures S6-S9).

@@

## Objective 3 - Application to real clinical PET data 
The third and final aim of this tutorial is to apply sequential BF testing to a real clinical PET setting. To this purpose, we used already collected data of patients with major depressive disorder and healthy control subjects examined with [11C]WAY100635, a radioligand, which binds to the serotonin 1A (5HT1A) receptor [@Parsey2010; @Chen2019]. From this data, we included 40 medication-free patients (mean Age 36.2 (12.9SD); 25 females) and 40 healthy controls (mean Age 37.1 (14.0SD), 25 females), using BP~F~ (specific binding over the free-fraction in plasma) as the primary outcome measure for assessing 5HT1A receptor availability in the brain. The 5HT1A receptor acts inhibitory on serotonergic neurons in the raphe nuclei. A high concentration of receptors in raphe will likely lead to lower transmitter release in serotonergic projection areas. Hence, we hypothesized that patients suffering from MDD show higher 5HT1A receptor availability in the raphe nuclei compared to healthy controls. Here we apply sequential BF testing to examine whether this hypothesis gains enough support to stop data collection before 40 subjects per group are reached.

@@

When applying sequential testing, the BF passes the threshold of 5 in support of H~1~, after the inclusion of 27 subjects/group (Figure \ref{fig_seqWAY}). Hence, had sequential testing been applied in collection of [11C]WAY patient-control data, it would have been possible to stop the recruitment at a total N of 54 instead of 80, saving 33% in terms of expense and radioactivity exposure. Assuming that a PET examination costs 5000 USD/Euro, that would amount to saving 130,000 USD/Euro in total.

@@

#### Alternative sequential procedures
There are sequential testing methods other than BF that can be used to stop the data collection early, and that offer exact control over error rates. For the NHST framework, a viable alternative is to apply a so-called group-sequential NHST design [@Schulz2005]. In fact, when calibrating the BF threshold (using a "default" Cauchy-distribution for describing H~1~) to a 5% false positive rate when stopping only for H~1~ (see  Objective 1), the sequential BF procedure will be identical to performing corrected sequential NHST t-tests (supplementary Figure XXX). For NHST, there are several different efficient strategies for doing sequential testing to stop a study as soon as H~0~ can be rejected [@Proschan2006; @Lakens2014]. One alternative is to use "alpha-spending functions", where the significance threshold changes for each sequential test, making it harder to reject H~0~ at an early stage of the study, but easier later on (or vice versa), while still keeping the total false positive rate below 5% [@Demets1994; @Selwyn2004]. There are also methods for performing so called "futility testing", with which the researchers can stop a study early if it is deemed to  unlikely that a significant finding will be achieved later on [@Lachin2005]. Addressing such sequential-NHST procedures for application in clinical PET research are however outside the scope of this tutorial.

#### Evidence in favor of H~0~
One advantage of sequential BF testing is that a PET study can also be stopped early when there is no effect in the population, i.e., when H~0~ is true. If the researcher wants to use a sequential BF to assess support for both H~1~ and H~0~, we recommend using a one-tailed default BF t-test. This entails that the researcher must make an a priori prediction on the direction of the effect (e.g., patients have higher binding than control subjects, or vice versa). Without using one-sided default BF t-test, it will be difficult (or even impossible) to stop a study early in support for H~0~, given the sample sizes used in typical PET studies (N<100 subjects/group) while still using a threshold that is high enough to keep the rate of false evidence at reasonable levels (see Supplementary Figure S11).

@@

#### General considerations
By using a directional hypothesis (one-sided test) when applying BF sequential testing it is possible to stop a study at low sample sizes, when the null-hypothesis is supported. If the researcher wishes to allow stopping for H~0~, but uses a two-tailed default BF t-test instead,the average N will be much higher when using a reasonable BF decision threshold (>3, see Supplementary Figure S11).

@@

Increasing the width of the Cauchy can be desirable in e.g., a paired study design, where the variance can be expected to be low, but the difference in raw scores is assumed to be similar to that from a cross-sectional design. The reason for this is that the same change in raw score will correspond to a much larger standardized effect size (see Table 2), and it is therefore sensible to specify an H~1~ which predicts a more extreme difference.  The lower rate of false positives that results from using a higher *r* (e.g., 1 instead of 0.707) can be utilized by starting the sequential BF testing earlier than at 12 subjects/group. As a result, the average sample size needed to reach a decision can be further decreased (see Supplementary Figure S10).

>>>>

Figure \ref{fig_k_at_5fpr}A and 3B show the estimated false positive rate when applying sequential BF testing in the case in which there is no difference between the two groups. Three different stopping thresholds have been used (BF > 3, 4 and 6, respectively). When the maximum allowed sample size increases, the false positive rate goes up. This is because the longer the sequential testing can go on, the more decisions are being made. Some of these decisions will be wrong, meaning that the researcher will claim support for an effect, when in fact there is no effect in the population. Eventuelly, when N becomes high enough, the rate of false positives will reach an asymptote (see Supplementary Figure S1). But with a maximum sample size above 30 subjects/group, it is not possible to keep false positives below the 5% using a decision threshold of 4. Instead, if the researchers wish to control the false positives at the 5% level, one option is to increase the decision threshold. Figure \ref{fig_k_at_5fpr}C and 3D show the estimated decision threshold that would be needed to keep the false positive rate at 5% when using sequential testing with the default BF t-test, while still starting at 12 subjects/group and checking the BF after the collection of each patient-control pair.

@@

Figure \ref{fig_powerPhase1} shows the results from the simulations when Nmax=30 subjects/group, using a decision threshold just above 4. In the upper panel the rate of positive studies at different population effects is visualised. The curves cross the y-axis at D=0 (i.e., the false positive rate), 5% for both BF seq and the conventional fixed N test. As can be seen, had the researcher instead applied a fixed N approach (at 30 subjects/group), they would have increased the power to detect the population effect by no more than ~10% for any effect size compared to using sequential testing. Note that when calibrating the BF threshold to correspond to a 5% false positive rate, and stopping only for H~1~, the BF procedure will show identical results to performing sequential NHST t-tests where the significance level has been corrected (see Supplementary Figure S2).

@@

Simulation results for different choices of Nmax (15, 20, 50 and 100 subjects/group) are presented in supplementary information (Supplementary Figures S3-S6).

@@

The same results but for different Nmax (15, 20, 50 and 100 subjects/group) are presented in supplementary information (Supplementary Figures S7-S10).

@@

## Objective 3 - Application to real clinical PET data 
The third and final aim of this tutorial is to apply sequential BF testing to a real clinical PET setting. To this purpose, we used already collected data of patients with major depressive disorder and healthy control subjects examined with [$^{11}$C]WAY100635, a radioligand, which binds to the serotonin 1A (5HT1A) receptor [@Parsey2010; @Chen2019]. From this data, we included 40 medication-free patients (mean Age 36.2 (12.9SD); 25 females) and 40 healthy controls (mean Age 37.1 (14.0SD), 25 females), using BP~F~ (specific binding over the free-fraction in plasma) as the primary outcome measure for assessing 5HT1A receptor availability in the brain. The 5HT1A receptor acts inhibitory on serotonergic neurons in the raphe nuclei. A high concentration of receptors in raphe will likely lead to lower transmitter release in serotonergic projection areas. Hence, we hypothesized that patients suffering from MDD show higher 5HT1A receptor availability in the raphe nuclei compared to healthy controls. Here we apply sequential BF testing to examine whether this hypothesis gains enough support to stop data collection before 40 subjects per group are reached.

@@

When applying sequential testing, the BF passes the threshold of 5 in support of H~1~, after the inclusion of 27 subjects/group (Figure \ref{fig_seqWAY}). Hence, had sequential testing been applied in collection of [$^{11}$C]WAY100635 patient-control data, it would have been possible to stop the recruitment at a total N of 54 instead of 80, saving 33% in terms of expense and radioactivity exposure. Assuming that a PET examination costs 5000 USD/Euro, that would amount to saving 130,000 USD/Euro in total.

@@

#### Alternative sequential procedures
There are sequential testing methods other than BF that can be used to stop the data collection early, and that offer exact control over error rates. For the NHST framework, a viable alternative is to apply a so-called group-sequential NHST design [@Schulz2005]. In fact, when calibrating the BF threshold (using a "default" Cauchy-distribution for describing H~1~) to a 5% false positive rate when stopping only for H~1~ (see  Objective 1), the sequential BF procedure will be identical to performing corrected sequential NHST t-tests (Supplementary Figure S2). For NHST, there are several different efficient strategies for doing sequential testing to stop a study as soon as H~0~ can be rejected [@Proschan2006; @Lakens2014]. One alternative is to use "alpha-spending functions", where the significance threshold changes for each sequential test, making it harder to reject H~0~ at an early stage of the study, but easier later on (or vice versa), while still keeping the total false positive rate below 5% [@Demets1994; @Selwyn2004]. There are also methods for performing so called "futility testing", with which the researchers can stop a study early if it is deemed to  unlikely that a significant finding will be achieved later on [@Lachin2005]. Addressing such sequential-NHST procedures for application in clinical PET research are however outside the scope of this tutorial.

#### Evidence in favor of H~0~
One advantage of sequential BF testing is that a PET study can also be stopped early when there is no effect in the population, i.e., when H~0~ is true. If the researcher wants to use a sequential BF to assess support for both H~1~ and H~0~, we recommend using a one-tailed default BF t-test. This entails that the researcher must make an a priori prediction on the direction of the effect (e.g., patients have higher binding than control subjects, or vice versa). Without using one-sided default BF t-test, it will be difficult (or even impossible) to stop a study early in support for H~0~, given the sample sizes used in typical PET studies (N<100 subjects/group) while still using a threshold that is high enough to keep the rate of false evidence at reasonable levels (see Supplementary Figure S12).

@@

#### General considerations
By using a directional hypothesis (one-sided test) when applying BF sequential testing it is possible to stop a study at low sample sizes, when the null-hypothesis is supported. If the researcher wishes to allow stopping for H~0~, but uses a two-tailed default BF t-test instead,the average N will be much higher when using a reasonable BF decision threshold (>3, see Supplementary Figure S12).

@@

Increasing the width of the Cauchy can be desirable in e.g., a paired study design, where the variance can be expected to be low, but the difference in raw scores is assumed to be similar to that from a cross-sectional design. The reason for this is that the same change in raw score will correspond to a much larger standardized effect size (see Table 2), and it is therefore sensible to specify an H~1~ which predicts a more extreme difference.  The lower rate of false positives that results from using a higher *r* (e.g., 1 instead of 0.707) can be utilized by starting the sequential BF testing earlier than at 12 subjects/group. As a result, the average sample size needed to reach a decision can be further decreased (see Supplementary Figure S11).
